Type-Safe Modular Parsing

### 2016-12-17 paper/sections/OpenRecursion.tex

where function \lstinline{f} has a by-name parameter of type \lstinline{"=>} \lstinline{A"}.
Here we illustrate by the famous Fibonacci example. Traditionally it can be implemented in Scala using pattern-matching:
\lstinputlisting[linerange=34-40]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB
An alternative implementation using open recursion is as follows:
\lstinputlisting[linerange=44-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
where \lstinline{self} is the explicit self-reference, namely the delegator. Functions like \lstinline{fib2} which takes
an explicit self-reference as its parameter are called \textit{generators} (by Ralf Hinze).
If we close the recursion on \lstinline{fib2} only, the evaluation of \lstinline{fix(fib2)(2)}, namely the above \lstinline{x2} is processed as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix(fib2)(2)
= lazy val a = fib2(a); a.apply(2)
= lazy val a = fib2(a); fib2(a)(2)
= ... (the third case in fib2)
= lazy val a = fib2(a); a.apply(1) + a.apply(0)
= lazy val a = fib2(a); fib2(a)(1) + fib2(a)(0)
= ... (the first two cases in fib2)
= lazy val a = fib2(a); 1 + 0
= 1
\end{lstlisting}
which behaves similarly as \lstinline{fib(2)}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.

>>>>

where function \lstinline{f} has a by-name parameter of type \lstinline{"=>} \lstinline{A"}.
Here we illustrate by the famous Fibonacci example. Traditionally it can be implemented in Scala using pattern-matching:
\lstinputlisting[linerange=34-40]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB

An alternative implementation using open recursion is as follows:
\lstinputlisting[linerange=44-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
where \lstinline{self} is the explicit self-reference, namely the delegator. Functions like \lstinline{fib2} which takes
an explicit self-reference as its parameter are called \textit{generators} (by Ralf Hinze).
If we close the recursion on \lstinline{fib2} only, the evaluation of \lstinline{fix(fib2)(2)}, namely the above \lstinline{x2} is processed as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix(fib2)(2)
= lazy val a = fib2(a); a.apply(2)
= lazy val a = fib2(a); fib2(a)(2)
= ... (the third case in fib2)
= lazy val a = fib2(a); a.apply(1) + a.apply(0)
= lazy val a = fib2(a); fib2(a)(1) + fib2(a)(0)
= ... (the first two cases in fib2)
= lazy val a = fib2(a); 1 + 0
= 1
\end{lstlisting}
which behaves similarly as \lstinline{fib(2)}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.

### 2016-12-17 paper/sections/Overview.tex

@@

Another issue of left-recursion elimination is that it requires extra bookkeeping work to retain the original semantic. For example, the expression $1+2-3$ is parsed as $(1+2)-3$ in the left-recursive grammar, but after rewrite the result is $1((+2)-3))$. The parse tree must be transformed to recover its structure.

\huang{done, the example may be not very good, but i haven't found a better one}\bruno{Illustrate with a concrete example: show a simple
  left-recursive grammar; then do left-recursion elimination. Then add
a new case into the original left-recursive grammar; and show that
a very different grammar is obtained after left-recursion elimination.}

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memorization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they supports left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

\paragraph{Client Code} Since the type of parsing functions are changed, we can not apply them directly. The client code below shows how to use the new parser. The \inlinecode{fix} function, which is the the standard \textit{fixpoint combinator}, is used to `close' the open recursion.

>>>>

\bruno{Write a summary for this section}
\bruno{Fill in all the missing references}

@@

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1+2-3$ is parsed as $(1+2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((+2)-3))$. The parse tree
must be transformed to recover its structure.

\huang{done, the example may be not very good, but i haven't found a
  better one}
\bruno{This example does not appear to be problematic: adding ``-''
  does not affect the rest of the grammar. Isn't the problem that
  sometimes adding a new case to the grammar, requires changing other 
parts? Otherwise I don't see why left-recursion is a problem, and we
cannot simply claim that it is a problem. You have to make a better
effort to find a problematic grammar.
}

@@

\bruno{The example is good, I think but the explanation is not. 
You want to say that when you know
  the full grammar, you can figure out where ``try'' is
  needed. Without the full grammar you'd need assume the worst
  case. Make an effort to make your explanation cristal clear!
I think you want to miss the abstract explanation that you give first,
with the explanation about the concrete example. 
Start with ``The need for backtracking is also problematic
in a modular setting. For example, suppose ...'' and synchronize 
the abstract explanation and the explanation for the example.
}

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

\paragraph{Client Code} Since the type of parsing functions are
changed, we can not apply them directly. The client code below shows
how to use the new parser. The \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. \bruno{Too much detail is being given here about open
  recursion. In Section 2 we should talk only about \emph{how} to use 
open recursion, and not how open recursion is implemented. We 
can summarize, for example the API related to open recursion. 
Section 4 is where the implementation and detailed explanation 
about open recursion should come.
}

### 2016-12-17 paper/sections/PackratParsers.tex

\paragraph{An Example}
A Packrat parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the representation\bruno{the
  representation of what?}. The code below illustrates
Scala's Packrat parsing library.

@@

\paragraph{Longest Match Composition} The Packrat parsing library
contains an alternation combinator, which does \emph{longest match
  composition}. This combinator has a different semantics from
\lstinline{|}.
Suppose that we are composing a list of parsers using alternative
\lstinline{|}, in an order. A parser in the front might be successful
in parsing a substring of the input, but then it unexpectedly stops
the parsing and discards the rest, while another parser might be able
to parse the whole input. In that case, using \lstinline{|||} we no
longer need to worry about the order of composition, which potentially
encourages its extensibility. A concrete example will be illustrated
for this later. \bruno{Please show example here. }

>>>>

\paragraph{An Example}
A Packrat parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces.
%\bruno{the representation of what?}
The code below illustrates
Scala's Packrat parsing library.

@@

\paragraph{Longest Match Composition} The Packrat parsing library
contains an alternation combinator, which does \emph{longest match
  composition}. This combinator has a different semantics from
\lstinline{|}.
Suppose that we are composing a list of parsers using alternative
\lstinline{|}, in an order. A parser in the front might be successful
in parsing a substring of the input, but then it unexpectedly stops
the parsing and discards the rest, while another parser might be able
to parse the whole input. In that case, using \lstinline{|||} we no
longer need to worry about the order of composition, which potentially
encourages its extensibility. As a concrete example, consider the grammar
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int>
\end{grammar}
which might be implemented in Packrat by \lstinline{pExpr = pInt | pAdd}, where \lstinline{pInt}
and \lstinline{pAdd} handle the two cases respectively. It does not work as expected even for input \lstinline{"1 + 2"},
as \lstinline{pInt}, which is applied first, parses \lstinline{1} and returns. Whereas \lstinline{pExpr = pInt ||| pAdd}
can successfully parse the whole string. Without the longest match composition, adjusting the order of alternatives is possible
yet inconvenient with extensible parsing.

%\bruno{Please show example here. }

### 2016-12-17 paper/sections/RelatedWork.tex

Our work integrates several components including extensible parsing, parser combinators and modular datatypes. There has been a great amount of related papers
on those hot topics, of which some inspired us of this paper and encourages us for more exploration. This section will try to lead a discussion on what difference we have made.

\paragraph*{Extensible Parsing} Extensible parsing is achieved in many different areas, of which parser generators are a mainstream area specially designed for modular syntax and parsing. Many parser generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support modular grammars, more specifically, they allow users to create new modules where new non-terminals and production rules can be introduced, some can even override existing rules in the old grammar modules. For instance, \textit{Rats!}~\cite{Grimm2006}
constructs its own module system for the collection of grammars, while NOA~\cite{Gouseti2014} uses Java annotation processing to gather information together. Those parser combinators focus on the \textit{syntactic extensibility} of grammars, and rely on a whole compilation to generate a global parser, even with a slight modification. Some of them may statically check the correctness and unambiguity of grammars, but separate compilation and modular type-checking remain unsolved, together with the issue of performance.

Besides, macro systems like C preprocessor, C++ templates and Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming techniques are a similar area aiming at syntactic extensibility as well, which sacrifices type safety. SugarJ~\cite{Erdweg2011} is another well-known tool that conveniently introduces syntactic sugars in Java programming by library imports. Composition of syntactic sugars is easy for users, whereas it requires many rounds of parsing and adaption, which highly affects efficiency of compilation, moreover, the implementation was based on SDF~\cite{Heering1989} and Stratego~\cite{Visser2001}, which focused little on separate compilation. Extensible compilers like JastAdd~\cite{Ekman2007} and Polyglot~\cite{Nystrom2003}, however, are somehow more ambitious, but they involve extensible parsing mostly by parser generators as well, and they focus more on the extensions to a host language. Our library is designed for modular language parsers in a type-safe way, with flexible language composition. Although overriding existing production rules is tricky with our approach, we could perhaps make use of embedded domain-specific languages on top of parser combinators and transformations for overriding, but it is anyway an orthogonal issue. \haoyuan{???} \haoyuan{BTW what about Racket?} \haoyuan{what about metafront? it is a macro system but does it have type safety?} \haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible syntax" proposes a system for extensible syntax, where users write EDSLs in their language with concrete syntax. Users can write rules for type-based disambiguation. But separate compilation is again not mentioned. Shall we mention that thesis?} \haoyuan{attribute grammars?}

On the other hand, extensible parsing algorithms are another area that indeed relates to separate compilation. Specifically, \cite{Bravenboer2009} introduces \textit{parse table composition}, in which paper grammars are compiled to the generation of parse tables, which are DFAs or NFAs, later they can be composed by an algorithm, so as to provide separate compilation for parsing. Nevertheless, the generation of parse tables is rather expensive, and furthermore, our approach supports separate compilation as well as modular type-checking, and the idea is not restricted to Scala but applicable to many functional languages, on whose type system the safety of modular parsing can rely. The extensibility of parsing in our approach is further available at language composition and lexical level.
\haoyuan{I mentioned one shortcoming of our approach here, which is we cannot override existing production rules; another one is that we do not have explicit correspondence/relationship between abstract syntax and the parser.}

\paragraph*{Parser Combinators} Since~\cite{burge1975} firstly introduced parser generators, and after Wadler discussed more details on backtracking in~\cite{Wadler1985}, parser combinators have been more and more popularly developed and used in the research area of parsing. Among them many parsing libraries produces recursive descent parsers by introducing functional, or more specifically, monadic parser combinators with a foundation on~\cite{nott237}. Related work includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell for context-sensitive grammars with infinite lookahead. Nevertheless, left-recursion is known as a big issue in recursive descent parsers, in which case programmers using libraries like Parsec have to manually eliminate left-recursion in the grammar, which is cumbersome, but also it distorts the shape of its old grammar, restricting the modularity of parsing to extreme extent.
There are certainly some approaches to afford the limitation of parser combinators, for example, we have conducted our previous experiment in Haskell, by using open recursion, MRM~\cite{Oliveira2015} and Parsec. To support direct left-recursive grammars, we made use of the state monad in Parsec to ensure that a left-recursive production rule will not be applied successively in parsing. It turned out to be successful, yet introduced complexity in programming, hence we believe that the parsing library is obligated for left-recursion.

Some more recent papers~\cite{Ford2002,Might2011,Frost2008} \haoyuan{Pracitical, general parser combinators?} proposed a series of novel parsing techniques, moreover, they did put an eye on the issue of left-recursion. As we explained before, we have selected Packrat Parsing for our prototype implementation, as the combinators are convenient to use in Scala, which is a platform where functional laziness and Object Algebras can perfectly coexist. Also, its support for direct left-recursion is already helpful for designing real-world parsers practically, though~\cite{warth2008} further demonstrated that general left-recursion is supported from the theoretical point of view. We have also mentioned that the components used in our library can have alternatives, for instance, a different set of parser combinators may support ambiguous and left-recursive grammars, or be applicable to a different subset of context-free grammars or more, or even lead us to another level of performance.

Another important issue in previous papers, to the best of our knowledge, is that none of them deal with modular datatypes or abstract syntax trees. The modularity of data structures is yet necessary for modular parsing that supports separate compilation and modular type-checking. In our library, Object Algebras have been adopted for extensible ASTs, together with the behaviors on them.

\paragraph*{Modular Datatypes} There has been lots of related work on building modular data structures, especially in some functional languages with non-extensible datatypes like Haskell. For instance, the famous DTC paper~\cite{swierstra2008} represents modular ASTs using co-products of every two functors, which forms a binary tree structure. It supports extensible datatypes, as well as the operations on them by implementing instances of functors. Later Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is similar to DTC but adopts a list of functors in type-level lists instead. As mentioned above, we have conducted our early experiment with MRM for modular datatypes, whereas the complexity of library implementation was comparably high. Other related work includes~\cite{Bahr2014}, where data types are flexible with composition as well as decomposition.

@@

Another interesting observation, as we have mentioned in~\ref{subsec:objectalgebras}, is that transformation algebras are frequently used in parsing, as concrete syntax may not always coincide with its abstract syntax, for instance, the bruijn indices in lambda calculus, as well as other syntactic sugars. In the Shy framework~\cite{Zhang2015}, the pattern of transformations is captured into a template for Java annotation processing, and most of boilerplate code is automatically generated, hence users only need to override a few interesting cases, conveniently. The idea of overriding existing cases was also proposed in MRM. It is potentially useful to our framework.

>>>>

Our work integrates several components including extensible parsing,
parser combinators and extensibility techniques. There has been a
great amount of related papers on those hot topics, of which some
inspired us of this paper and encourages us for more exploration. This
section will try to lead a discussion on what difference we have made.

\paragraph*{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, of which parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with the issue of performance.
\bruno{what performance issue?}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility. 
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which focused little on separate
compilation\bruno{focused little or did not have it at all?}. 
Extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover 
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition. Although overriding existing production
rules is tricky with our approach, we could perhaps make use of
embedded domain-specific languages on top of parser combinators and
transformations for overriding, but it is anyway an orthogonal issue.
\bruno{Not sure that I understand the last sentence}
\haoyuan{???} \haoyuan{BTW what about Racket?} \bruno{Have you read 
Languages as Libraries? That is probably an important reference, which
I think we should cite.}\haoyuan{what about
  metafront? it is a macro system but does it have type safety?}
\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
  syntax" proposes a system for extensible syntax, where users write
  EDSLs in their language with concrete syntax. Users can write rules
  for type-based disambiguation. But separate compilation is again not
  mentioned. Shall we mention that thesis?} \haoyuan{attribute
  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, in which paper grammars\bruno{what
  is a paper grammar?} are compiled
to the generation of parse tables\bruno{I don't understand the last
  part; please rephrase. Also please break the sentence into multple
  sentences. Generally speaking go over the text you wrote and watch
  out for long sentences. You have a tendency to write very long
  sentences. }, which are DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive\bruno{expensive in terms of what? why is it expensive?}. 
Furthermore, \name supports \emph{both} separate compilation
as well as \emph{modular type-checking}. The extensibility of
parsing in our approach is further available at language composition
and lexical level. \haoyuan{I mentioned one shortcoming of our
  approach here, which is we cannot override existing production
  rules; another one is that we do not have explicit
  correspondence/relationship between abstract syntax and the parser.}

\paragraph*{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
left-recursion is known as a big issue in recursive descent parsers,
in which case programmers using libraries like Parsec have to manually
eliminate left-recursion in the grammar, which is cumbersome, but also
it distorts the shape of its old grammar, restricting the modularity
of parsing to extreme extent. It is possible to avoid some of  
 the limitations of parser combinators. 
\bruno{I'm not sure that th following text should be here. It could 
be in the paper, but somewhere else.}
For example, we have
conducted our previous experiment in Haskell, by using open recursion,
MRM~\cite{Oliveira2015} and Parsec. To support direct left-recursive
grammars, we made use of the state monad in Parsec to ensure that a
left-recursive production rule will not be applied successively in
parsing. It turned out to be successful, yet introduced complexity in
programming, hence we believe that the parsing library is obligated
for left-recursion.

Some more recent papers~\cite{Ford2002,Might2011,Frost2008}
\haoyuan{Pracitical, general parser combinators?} proposed a series of
novel parsing techniques that addressed the issue
of left-recursion. We have selected Packrat
Parsing for our prototype implementation, as the combinators are
convenient to use in Scala, which is a platform where functional
laziness and Object Algebras can perfectly coexist. Also, its support
for direct left-recursion is already helpful for designing real-world
parsers practically, though~\cite{warth2008} further demonstrated that
general left-recursion is supported from the theoretical point of
view. We have also mentioned that the components used in our library
can have alternatives, for instance, a different set of parser
combinators may support ambiguous and left-recursive grammars, or be
applicable to a different subset of context-free grammars or more, or
even lead us to another level of performance.

To the best of our knowledge, no previous work on parser combinators 
deals with semantic modularity and
extensibility of parsing. The modularity of abstract syntax trees is
necessary for modular parsing that supports separate compilation and
modular type-checking. In our library, Object Algebras have been
adopted for extensible ASTs, together with the behaviors on them.

\paragraph*{Extensibility} There has been alot of work on building
modular programs, in both functional and object-oriented programming .
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. It supports extensible datatypes, as well as
the operations on them by implementing instances of functors. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. As mentioned above, we have conducted our early experiment
with MRM for modular datatypes, whereas the complexity of library
implementation was comparably high. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.

@@

Another interesting observation, as we have mentioned
in~\ref{subsec:objectalgebras}, is that transformation algebras are
frequently used in parsing, as concrete syntax may not always coincide
with its abstract syntax, for instance, the bruijn indices in lambda
calculus, as well as other syntactic sugars. In the Shy
framework~\cite{Zhang2015}, the pattern of transformations is captured
into a template for Java annotation processing, and most of
boilerplate code is automatically generated, hence users only need to
override a few interesting cases, conveniently. The idea of overriding
existing cases was also proposed in MRM. It is potentially useful to
our framework.\bruno{The discussion on Extensibility can be
  considerably shorter. We just need to point out that there are many
  techniques for extensibility (without discussing alot of details),
  but most of that work does not address the problem of 
\emph{producing extensible ASTs} (rather than consuming those. Then
you can talk about Baht's work, and have the discussion on
transformations. Generally speaking, go over your text, polish it, and
trim it. You are a bit too ``chatty'' (that is you talk too much): go
straight to the point.}
